The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-06-17 paper/ms.tex
includes plots for the analytic solutions
going through the paper

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. We model a galaxy as a sphere with
an homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with maximum
velocities in the range $0-300$ \kms and neutral hydrogen optical
depths in the range $\tau_{\rm H}=10^{5}-10^{7}$. We consider cases 
of a single central \lya source, and a uniform distribution of sources. 
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle. Observations with a line of sight
parallel to the rotation axis yield line morphologies similar to the
static case. The greatest difference is observed for lines of sight
perpendicular to the rotation axis. In this case the intensity at the line
center increases and the line width increase with rotational velocity.  For
homogeneously distributed sources the line becomes single peaked  at
rotational velocities larger than half the line width in the static
case. Notably, we find that rotation does not not induce any
significant spatial anisotropy in the integrated line flux nor in the escape
fraction.  However, we do find in the homogeneous models a clear
increase of the escape fraction with increasing rotational velocity.
\end{abstract}

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f2.pdf}
\end{center}
\caption{
2D histogram showing the number of photons that escape with frecuency
$x$ forming an angle $\theta$ (parameterized as $|\cos\theta|$) with the
rotation axis. The rotational velocity ($0,100,200,300$\kms)
increases from left to right and the optical depth ($10^5$, $10^6$,
$10^7$) from top to bottom. The \ly photons are initialized at the
center of the sphere.
\label{fig:CentralSpec} }   
\end{figure*}

@@

Here we briefly describe the relevant points for the two radiative
transfer codes we have used. For a detailed description we refer the
reader to the original papers \cite{CLARA,DijkstraKramer}.

The codes follow the individual scatterings of \ly photons as they
travel through a 3D distribution of neutral Hydrogen. At each
scattering the frequency of the photon (in the laboratory frame) and
its direction of propagation change. This change in frequency is due
to the peculiar velocities of the Hydrogen atom that absorbed and
re-emitted the photon. If dust is present, the photon can interact
either with a Hydrogen atom or dust grain. In the case of a dust
interaction the photon can be either absorbed or scattered, this
probability is encoded in the dust albedo, $A$, which we chose to be
$1/2$. In order to obtain accurate values for the escape fraction of
photons in the presence of dust, we do not use any accelerating
mechanism in the radiative transfer. Once the photons escape the gas
distribution we store their direction at their direction of
propagation and frequency at their last scattering.

The photons are thus emitted in some region of the gas distribution
and follow a random walk in space and frequency until they escape the
gas distribution or are absorbed by a dust grain. The initialization
process for the \ly photons has to specify its position, frequency and
direction of propagation. In our case we select the initial frequency
to be exactly the \ly rest-frame frequency ($x=0$) and the direction of
propagation to be random following an flat probability distribution
over the sphere. A different initization that uses a Gaussian with a
velocity with equal to the thermal velocity, $12.8$\kms in our case,
should not introduce noticeable change given that our rotational
velocities span the range $100$-$300$\kms and the lines have velocity
widths on the order of $100$-$500$\kms.

The gas is completely defined by its geometry (i.e. sphere or slab),
temperature $T$, Hydrogen optical depth $\tau_{\rm H}$, dust optical
depth $\tau_{\rm a}$ and the bulk velocity field $\vec{v}$. Our codes
treat the gas as homogeneous in density ($\tau_{\rm H}$, $\tau_{\rm
  a}$) and temperature.

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies. For each galaxy
we vary at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, the hydrogen optical depth $\tau_{H}$,
the dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. There are $60$ models initial combining all
variations of the input parameters. Table \ref{table:models}
summarizes the different parameters.

Additionally, we have used two independently developed Monte Carlo
codes \citep{CLARA,DijkstraKramer} to perform the calculations. The
results we report are robust in the sense that they are obtained by
both codes.

\begin{table}
\begin{center}
\begin{tabular}{llc}\hline\hline
Physical Parameter (units) & Symbol & Values\\\hline
Velocity (\kms) & $V_{\rm max}$&$0,\ 50,\ 100,\ 200,\ 300$\\
Hydrogen Optical Depth & $\tau_{H} $ & $10^{5},\ 10^{6},\ 10^{7}$\\
Dust Optical Depth & $\tau_{a}$ & $0$,$1$\\
Photons Distributions & & Central, Homogeneous\\\hline\hline
\end{tabular}
\caption{
  Summary of Physical Parameters of our Monte Carlo Simulations.} 
\label{table:models}
\end{center}
\end{table}

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.40\textwidth]{f10.pdf}
  \includegraphics[width=0.40\textwidth]{f11.pdf}
\end{center}
  \caption{Effects of rotation on the escape fraction. Left
    panel, escape fraction as a function of rotational velocity. All
    these models have $\tau_{a}=1$. The continuous (dashed) lines
    correspond to central (homogeneous) models. Right panel, escape
    fraction as a    function of the product $(a\tau_{\rm
      H})^{1/3}\tau_{A}$. The    analytic solution for the infinite
    slab is shown as a continuous line. Different lines correspond
    to different rotational velocities for the homogeneous models.     
    \label{fig:escape_fraction}} 
\end{figure*}

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=0.45\textwidth]{f12.pdf}
    \includegraphics[width=0.45\textwidth]{f13.pdf}
\end{center}
\caption{Logarithm of the average number of scatterings as function of
  $\mu$ (left panel) and the maximum rotational velocity $V_{\rm
    max}$. The left panel shows the behaviour for $\tau=10^{5}$ and
  $V_{max}=300$\kms as a function of $|\cos\theta|$; the independence of
  $N_{\rm scatt}$ with $\mu$ is present in all models. 
  In the right panel the continuous (dashed) lines represent an
  central (homogeneous) distribution of sources; there is a clear
  decrease in the number of scatterings for the homogenously
  distributed sources. 
\label{fig:Nscatt} }    
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{f14_1.pdf}
  \includegraphics[width=0.48\textwidth]{f14_2.pdf}
  \includegraphics[width=0.48\textwidth]{f14_3.pdf}
  \includegraphics[width=0.48\textwidth]{f14_4.pdf}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left panels 
      corresponds to the static case and right panels correspond to
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}. A results that stands out is in the upper
      right panel, showing that in homogeneously distributed sources a
      large number of photons can escape escape after a few
      scatterings, $N_{\rm scatt}<10$.}   
\end{figure*}

@@

%\begin{figure*}[h]
%\centerline{\includegraphics[width=85mm,angle=270]{spec.eps}}
%\caption[]{Comparison between analytic ({\it red solid lines}) \& Monte-Carlo ({\it blac%k histograms}) calculations of spectra. Left panel shows spectrum when the observer looks down the rotation axis. The right panel shows the spectra when the angle between the line-of-sight and rotation axis is ~45 degrees. Note that here $\mu \equiv \cos i$.}
%\label{fig:scheme}
%\end{figure*}

@@

%\begin{figure*}[h]
%\centerline{\includegraphics[width=85mm,angle=270]{spec2.eps}}
%\caption[]{Same as Fig~2. Right panel is now when $\cos i > 0.98$.}
%\label{fig:scheme}
%\end{figure*} 
%\end{document}

>>>>

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider cases of a single central \lya source, and a uniform
distribution of sources.  
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis the intensity at the line
center and the line width increase with rotational velocity.   
Along the same sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux nor in the escape fraction. 
We provide an analytical expression for the outgoing line intensity
that accounts for the velocity and viewing angle trends reported here.
\end{abstract}

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f2.pdf}
\end{center}
\caption{
2D histogram showing the number of photons that escape with frecuency
$x$ forming an angle $\theta$ (parameterized as $|\cos\theta|$) with the
rotation axis. 
The rotational velocity ($0,100,200,300$\kms) increases from left to
right and the optical depth ($10^5$, $10^6$, $10^7$) from top to
bottom. 
The \ly photons are initialized at the center of the sphere. 
Two main results can be read from this figure.
First, the line morphology depends on the viewing angle. 
Second, the line can become single peaked for high rotational
velocities.
\label{fig:CentralSpec} }   
\end{figure*}

@@

Here we briefly describe the relevant characteristics of the two
radiative transfer codes we have used.  
For a detailed description we refer the reader to the original papers
\cite{CLARA,DijkstraKramer}.

The codes follow the individual scatterings of \ly photons as they
travel through a 3D distribution of neutral Hydrogen. 
The frequency of the photon (in the laboratory frame) and
its direction of propagation change at every scattering. 
This change in frequency is due to the peculiar velocities of the
Hydrogen absorbing and re-emitting the photon. 
Once the photons escape the gas distribution we store their direction of
propagation and frequency at their last scattering.

The initialization process for the \ly photons especifies its
position, frequency and direction of propagation. 
We select the initial frequency to be exactly the \ly rest-frame
frequency in the restframe of the gas and the direction of propagation
to be random following an flat probability distribution over the sphere.  
It means that for photons emitted from the center of the sphere
$x_{\rm in}=0$, while photons emitted at some radii with a peculiar
velocity $\vec{v}$ have initial values $x_{in}$ depending on its direction of
propagation: $x_{\rm   in}=\vec{v}\cdot\hat{k}_{\rm in}/v_{\rm th}$.
We do not include the effect of turbulent velocities in the
initialization.
We neglect this given that the induced perturbation should be on the
close to the thermal velocity, $12.8$\kms, which is one order of
magnitude smaller than the velocity widths ($100$-$500$\kms) in the
static case.

If dust is present, the photon can interact either with a Hydrogen
atom or dust grain.
In the case of a dust interaction the photon can be either absorbed or
scattered.
This probability is encoded in the dust albedo, $A$, which we chose
to be $1/2$. 
In order to obtain accurate values for the escape fraction of
photons in the presence of dust, we do not use any accelerating
mechanism in the radiative transfer.

The codes treat the gas as homogeneous in density and temperature.  
This implies that the cas is completely defined by its geometry
(i.e. sphere or slab), temperature $T$, Hydrogen optical depth
$\tau_{\rm H}$, dust optical depth $\tau_{\rm a}$ and the bulk
velocity field $\vec{v}$.

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies. 
For each galaxy we vary at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, the hydrogen optical depth $\tau_{H}$,
the dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. 
In total there are $48$ different models combining all the possible
different variations in the input parameters.
Table \ref{table:models} lists the different parameters we used to
generate the models.

Additionally, we have used two independently developed Monte Carlo
codes \citep{CLARA,DijkstraKramer} to perform the calculations. 
The results and trends we report are observed in both codes.

\begin{table}
\begin{center}
\begin{tabular}{llc}\hline\hline
Physical Parameter (units) & Symbol & Values\\\hline
Velocity (\kms) & $V_{\rm max}$&$0,\ \ 100,\ 200,\ 300$\\
Hydrogen Optical Depth & $\tau_{H} $ & $10^{5},\ 10^{6},\ 10^{7}$\\
Dust Optical Depth & $\tau_{a}$ & $0$,$1$\\
Photons Distributions & & Central, Homogeneous\\\hline\hline
\end{tabular}
\caption{
  Summary of Physical Parameters of our Monte Carlo Simulations.} 
\label{table:models}
\end{center}
\end{table}

@@

%\begin{figure*}
%\begin{center}
%  \includegraphics[width=0.40\textwidth]{f10.pdf}
%  \includegraphics[width=0.40\textwidth]{f11.pdf}
%\end{center}
%  \caption{Effects of rotation on the escape fraction. Left
%    panel, escape fraction as a function of rotational velocity. All
%    these models have $\tau_{a}=1$. The continuous (dashed) lines
%    correspond to central (homogeneous) models. Right panel, escape
%    fraction as a    function of the product $(a\tau_{\rm
%      H})^{1/3}\tau_{A}$. The    analytic solution for the infinite
%    slab is shown as a continuous line. Different lines correspond
%    to different rotational velocities for the homogeneous models.     
%    \label{fig:escape_fraction}} 
%\end{figure*}

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{f14_1.pdf}
  \includegraphics[width=0.48\textwidth]{f14_2.pdf}
  \includegraphics[width=0.48\textwidth]{f14_3.pdf}
  \includegraphics[width=0.48\textwidth]{f14_4.pdf}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left panels 
      corresponds to the static case and right panels correspond to
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}. }
\end{figure*}

@@

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{vary_vel.pdf}
  \includegraphics[width=0.48\textwidth]{vary_angle.pdf}
\end{center}
\caption{
Comparison of the Monte Carlo results agains the analytic solution.
\label{fig:comparison} }   
\end{figure*}
